\begin{abstract}
  This thesis discusses algorithms for the uniform interpolation problem and presents their implementation for the following theories: (quantifier-free) equality with uninterpreted
  functions (EUF), unit two variable per 
  inequality (UTVPI), and theoretic aspects for the 
  combination of the two previous theories. 
  The uniform interpolation algorithms implemented in this thesis were 
  originally proposed in \cite{KAPUR2017}. 

  Refutational proof-based solutions are the usual approach 
  of many interpolation algorithms 
  \cite{10.1007/978-3-642-00768-2_34, mcmillan2011interpolants, 10.1007/978-3-540-24730-2_2}. 
  The approach taken in \cite{KAPUR2017} relies on quantifier-elimination heuristics 
  to construct a uniform interpolant using one of the two formulas involved 
  in the interpolation problem. The latter makes it 
  possible to study the complexity 
  of the algorithms obtained compared to refutational-based
  solution which rely on the efficiency of 
  SMT solvers.

  It is not always possible to find a
  uniform interpolant for every formula in 
  the combined theory of EUF and UTVPI 
  \cite{10.1007/978-3-030-51074-9_11}.
  Hence, the thesis work implements an
  algorithm for a subset of formulas in 
  the combined theory which
  the existence of uniform interpolants
  is guaranteed.
  Additionally, the thesis work implements
  a Nelson-Oppen interpolation framework 
  \cite{10.1007/11532231_26}
  to combine the uniform 
  interpolating
  algorithms in previous sections.
  
  The implementation uses Z3 
  \cite{10.1007/978-3-540-78800-3_24} 
  for parsing purposes and 
  satisfiability checking in 
  the combination component of the 
  thesis. Minor modifications were applied 
  to Z3's enode data structure 
  in order to label and distinguish formulas 
  efficiently (i.e. distinguish 
  A-part, B-part). The project can easily be 
  integrated into the Z3 solver 
  to extend its functionality for 
  verification purposes using the Z3 
  plug-in module. 
  
  The major results of the project are the 
  following:

  \begin{itemize}
    \item Implementation of Kapur's uniform interpolating
      algorithm for theories EUF and UTVPI.
    \item Modification and implementation 
      of the Phase III in Kapur's
      uniform interpolation algorithm for EUF. As
      a byproduct, an application in 
      membership testing in conjunction of Horn 
      clauses is obtained.
    \item Experimental evidence
      of uniform interpolants is provided as
      well as performance results of the 
      implemented systems.
    \item An partially sound uniform interpolation 
      algorithm for the combined theory of EUF and UTVPI
      is proposed and proven correct for a suitable 
      fragment of the aforementioned combined theory.
  \end{itemize}

  \clearpage %(required for 1-page abstract)
\end{abstract}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
