Theory combination techniques involve 
reusing the algorithms for verification 
problems of some of the theories involved 
by either purifying terms over the 
theories, or reducing the original problem 
into a base theory, or a combination of these two. 
Usually the approaches following the first approach 
aforementioned rely on a Nelson-Oppen
framework \cite{10.1007/11532231_26, 
10.1007/978-3-642-22119-4_1, 10.1145/2490253}

Despite possibly more efficient approaches using
model-based theory combination techniques 
\footnote{
  An interesting property of model-based theory combination 
  is that it is not needed to propagate disjunctions, even if
  the theories are not convex. The reason for this is because
  the \emph{disjunctiviness} of the problem is handled 
  \emph{lazily} by the
  SAT solver.
} \cite{10.1007/978-3-642-22119-4_1}, the latter requires 
operations which many SMT solvers do not provide 
proper API to implement these.
Hence, the approach used in the thesis work follows 
\cite{10.1007/11532231_26} since it does not require extensive
modification to the decision procedures
used. Many of the necessary modifications 
were implemented on top of Z3 and PicoSAT/TraceCheck.

The propagation of equalities and disjunction of equalities
requires the additional step to split these formulas into
the correct part of the interpolant pair. Among the major
contributions of \cite{10.1007/11532231_26} was the introduction
of the class of equality interpolanting theories, which
facilites the splitting of a propagated formula since
it is clear that A-local/B-local terms should be included in the
A-part/B-part of the interpolant pair respectively.

\begin{definition}
  A theory $\cal{T}$ is \emph{equality interpolanting}
  if for every $a$ (A-local term) and  $b$ (B-local term)
  and $\psi$ a formula in $\cal{T}$ such that
  $\psi \models_{\cal{T}} a = b$, then there exits a term $t$
  in $\cal{T}$ (called interpolanting term)
  such that $\psi \models_{\cal{T}} a = t$ and 
  $\psi \models_{\cal{T}} b = t$
\end{definition}

The relevance of the existance of the interpolating term for 
a deduced AB-mixed equality becomes relevant in the context 
of splitting a formula into a suitable A-part and B-part 
respectively. If t

Deciding where to include AB-common terms to either the 
A-part or the B-part of the interpolantion pair affects 
the final result since the interpolant will be \emph{closer} 
to the A-part or to the B-part respectively. The original
paper includes AB-common terms to the B-part. However,
the implementation work includes AB-common terms to the
A-part since the interpolantion algorithms implemented
focus the attention on this part.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
