\begin{dedication}
  To my family.
\end{dedication}

\begin{acknowledgments}
  \vspace{1.1in}

  First, I express immense gratitude to my
  advisor Prof. Deepak Kapur for his guidance 
  and support during the development of my 
  thesis work. His attention on intuitive 
  examples and work of high quality has
  reshaped my approach towards research in 
  general. 

  My appreciation for the committee members
  cannot be quantified in the standard
  model of numbers. I appreciate their patience 
  towards me.
  In particular, I would like to thank Dr. Mark
  Marron of Microsoft Research for discussions 
  regarding the performance
  of my implementation and how to show evidence
  that the implementation work can outperform
  other existing implementations if the 
  code were to be designed correctly. 
  Additionally, I would like to express 
  my appreciation towards Prof. Veroff, who kindly 
  pointed out how to improve several details 
  in the write-up, which were
  incorrect.

  I would like to acknowledge Prof. Naijun Zhang
  for his support and hospitality during a
  short academic staying at the Institute
  of Software Chinese Academy of Sciences, where
  I initially started working on the 
  implementation work using the C++ language.

  Finally, I would like to thank my family and
  friends whose unconditional love and support 
  where the motivation for me to keep working 
  on the project.

\end{acknowledgments}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
